--- 
title: NULISAseqR User Guide
site: bookdown::bookdown_site
documentclass: book
url: 'https://alamar-biosciences.github.io/NULISAseqR/'
github-repo: "Alamar-Biosciences/NULISAseqR"
description: "A comprehensive guide to analyzing NULISAseq proteomic data with the NULISAseqR R package"
---

# Welcome {-}

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6,
  cache = TRUE,
  comment = "#>"
)

# Load required libraries
suppressPackageStartupMessages({
  library(NULISAseqR)
  library(tidyverse)
})
```

This guide will walk you through analysis of NULISAseq proteomic data using R. 
<br>

If you have limited experience with R or prefer a graphical user interface, consider using the [NULISA Analysis Software (NAS)](#nulisa-analysis-software-nas) — a cloud-based platform designed for intuitive data processing, visualization, and analysis.


## What is NULISAseqR? {-}

**NULISAseqR** is an R package designed specifically for working with NULISAseq proteomic data. It provides:

- **Data Import**: Read XML files from NULISAseq multiplex proteomic panels
- **Quality Control**: Automated QC metrics and reporting
- **Visualization**: Heatmaps, PCA plots, volcano plots, boxplots, etc.
- **Statistical Analysis**: Differential abundance testing and longitudinal analysis
- **Outcome Modeling**: Linear models using single-protein abundance to predict an outcome

## Installation {-}

**New to R?** If you haven't installed R and RStudio yet, see the [Installing R and RStudio](#r-install) guide in the Additional Resources page.

### macOS Users {-}

**Important**: Before installing NULISAseqR on macOS, you need to install Xcode Command Line Tools:

1. Install Xcode from the App Store (if not already installed)
2. Open Terminal and run this code to accept the Xcode license agreement:
```bash
sudo xcodebuild -license accept
```

### Windows Users {-}

**Important**: Before installing NULISAseqR on Windows, you need to install Rtools which provides compilers and utilities to build packages from source:

1. Go to [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/){target="_blank"}
2. Download and install the appropriate Rtools version for your R installation
3. Follow the installation instructions on the website

### Installation Steps {-}

Follow these steps to install NULISAseqR and its dependencies:

```{r install, eval=FALSE}
# 1. Install devtools
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")

# 2. Install BiocManager for Bioconductor packages
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# 3. Install ComplexHeatmap from Bioconductor
BiocManager::install("ComplexHeatmap")

# 4. Install ggalt from CRAN snapshot
install.packages('ggalt', repos = "http://packagemanager.posit.co/cran/2025-08-02")

# 5. Install PCAtools (Alamar fork)
devtools::install_github('Alamar-Biosciences/PCAtools')

# 6. Install NULISAseqR
devtools::install_github('Alamar-Biosciences/NULISAseqR')

# 7. Install additional packages for data analysis
install.packages("tidyverse")
install.packages("table1")
install.packages("pROC")
```

> **Note**: If you encounter issues installing packages from source, you may need to restart R (Cmd/Ctrl + Shift + F10 in RStudio) between major installation steps.

<br>
Load the packages:

```{r load_packages, eval=FALSE}
library(NULISAseqR)
library(tidyverse)
```

## Getting Help {-}

- **Function help**: Type `?function_name` in the R console
- **GitHub Issues**: [Report bugs or request features](https://github.com/Alamar-Biosciences/NULISAseqR/issues){target="_blank"}


<hr>

<p style="text-align: center; font-size: 0.9em; color: #666;">
© `r format(Sys.Date(), "%Y")` Alamar Biosciences Bioinformatics Team<br>
Generated with <a href="https://github.com/Alamar-Biosciences/NULISAseqR" target="_blank"><strong>NULISAseqR</strong></a> version `r as.character(utils::packageVersion("NULISAseqR"))`<br>
Last updated: `r format(Sys.Date(), "%B %d, %Y")`
</p>
<p style="text-align: center;">
</p>